A Spectral Analysis of the Domain Decomposed Adjoint Neumann-Ulam Method

### 2012-11-29 doc/mandc2013/notes/spectral_theory.tex
adding new data to paper

In the case of domain decomposed linear operator equations, we can use
diffusion theory to estimate the optical thickness of a region and the
corresponding leakage fraction in terms of properties of the linear
operator and the discretization. To begin we must first calculate the
mean distance a Monte Carlo history will move in the grid by computing
the mean squared distance of its movement along the chord defined
across the domain of length $l$. After a single transition a history
will have moved a mean squared distance of:
\begin{equation}
  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
  \label{eq:step_1_length}
\end{equation}
where $h$ is the size of the grid along the chord and $n_s$ is the
number of grid elements a history will move on average every
transition. For our diffusion model problem, $n_s$ would equate to the
number of states in the $i$ (or $j$ as the problem is symmetric)
direction that a history could potentially move in a single transition
and is dependent on the stencil used for the discretization. After $k$
transitions in the random walk, the history will have moved a mean
squared distance of:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
  \label{eq:step_k_length}
\end{equation}
If our our chord is of length $l$ and there are $n_i$ grid points (or
states to which a history may transition along that chord), then $h =
l / n_i$ giving:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
  \label{eq:step_k_length_sub}
\end{equation}
From diffusion theory, we expect the average number of interactions
along the chord to be:
\begin{equation}
  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
  \label{eq:optical_thickness_1}
\end{equation}
where $d$ is the dimension of the problem and $\sqrt{\langle
  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
Carlo history in the domain. We can readily interperet $\tau$ to be
the \textit{effective optical thickness} of the domain of length
$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
  \label{eq:optical_thickness_2}
\end{equation}
which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
relation for the effective optical thickness:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s}
  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
  \label{eq:optical_thickness_3}
\end{equation}

Using this optical thickness, we can then complete the leakage
approximations by defining the bounds of $\tau \rightarrow 0, \Lambda
\rightarrow 1$ and $\tau \rightarrow \infty, \Lambda \rightarrow
\tau^{-1}$. We expect this behavior as optically thin domains will
communicate most of their histories while optically thick domains will
leak the fraction of histories that did not interact within. With
these bounds, we can then define the leakage fraction out of a domain
for the adjoint Neumann-Ulam method using the Wigner rational
approximation:
\begin{equation}
  \Lambda = \frac{1}{1+\tau}\:,
  \label{eq:wigner_domain_leakage}
\end{equation}
and using the mean-chord approximation:
\begin{equation}
  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
  \label{eq:wigner_domain_leakage}
\end{equation}

@@

To compute the leakage fraction numerically, over the global domain
\sn{1}{5} histories were sampled from a uniform source of strength
unity. At the start of a stage of histories, the number of histories
starting in the domain was computed and as the stage progressed, the
number of histories that exited the domain was tallied with the ratio
of the two numbers providing a numerical measure for the leakage
fraction. Figure~\ref{fig:measured_length} gives the domain leakage
measurements for the domain in the center of the global grid as well
as the analytic result computed by
Eq~(\ref{eq:analytic_domain_leakage}) as a function of the absorption
cross section. Again, we note good qualitative agreement between the
measured and analytic quantities.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{measured_leakage.png}
  \end{center}
  \caption{\textbf{Measured and analytic domain leakage as a function
      of the iteration matrix spectral radius.} \textit{To test the
      behavior with respect to domain size, $n_i=5$, $n_i=10$,
      $n_i=20$, and $n_i=100$ were used. The red data was computed
      numerically by a domain-decomposed adjoint Neumann-Ulam
      implementation while the black dashed data was generated by
      Eq~(\ref{eq:analytic_domain_leakage}).}}
  \label{fig:measured_leakage.}
\end{figure}

>>>>

In the case of domain decomposed linear operator equations, we can use
diffusion theory to estimate the optical thickness of a region and the
corresponding leakage fraction in terms of properties of the linear
operator and the discretization. To begin we must first calculate the
mean distance a Monte Carlo history will move in the grid by computing
the mean squared distance of its movement along the chord defined
across the domain of length $l$. After a single transition a history
will have moved a mean squared distance of:
\begin{equation}
  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
  \label{eq:step_1_length}
\end{equation}
where $h$ is the size of the grid along the chord and $n_s$ is the
number of grid elements a history will move on average every
transition. For our diffusion model problem, $n_s$ would equate to the
number of states in the $i$ (or $j$ as the problem is symmetric)
direction that a history could potentially move in a single transition
and is dependent on the stencil used for the discretization. After $k$
transitions in the random walk, the history will have moved a mean
squared distance of:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
  \label{eq:step_k_length}
\end{equation}
If our our chord is of length $l$ and there are $n_i$ grid points (or
states to which a history may transition) along that chord, then $h =
l / n_i$ giving:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
  \label{eq:step_k_length_sub}
\end{equation}
From diffusion theory, we expect the average number of interactions
along the chord to be:
\begin{equation}
  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
  \label{eq:optical_thickness_1}
\end{equation}
where $d$ is the dimension of the problem and $\sqrt{\langle
  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
Carlo history in the domain. We can readily interperet $\tau$ to be
the \textit{effective optical thickness} of the domain of length
$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
  \label{eq:optical_thickness_2}
\end{equation}
which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
relation for the effective optical thickness:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s}
  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
  \label{eq:optical_thickness_3}
\end{equation}

Using this optical thickness, we can then complete the leakage
approximations by defining the bounds of $\tau \rightarrow 0, \Lambda
\rightarrow 1$ and $\tau \rightarrow \infty, \Lambda \rightarrow
\tau^{-1}$. We expect this behavior as optically thin domains will
communicate most of their histories while optically thick domains will
leak the fraction of histories that did not interact within. With
these bounds, we can then define the leakage fraction out of a domain
for the adjoint Neumann-Ulam method using the Wigner rational
approximation:
\begin{equation}
  \Lambda = \frac{1}{1+\tau}\:,
  \label{eq:wigner_domain_leakage}
\end{equation}
and using the mean-chord approximation:
\begin{equation}
  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
  \label{eq:mean_chord_domain_leakage}
\end{equation}

@@

To compute the leakage fraction numerically, over the global domain
\sn{1}{5} histories were sampled from a uniform source of strength
unity. At the start of a stage of histories, the number of histories
starting in the domain was computed and as the stage progressed, the
number of histories that exited the domain was tallied with the ratio
of the two numbers providing a numerical measure for the leakage
fraction. Figure~\ref{fig:measured_length} gives the domain leakage
measurements for the domain in the center of the global grid as well
as the analytic result computed by
Eq~(\ref{eq:analytic_domain_leakage}) as a function of the absorption
cross section. Again, we note good qualitative agreement between the
measured and analytic quantities.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{leakage_variation.png}
  \end{center}
  \caption{\textbf{Measured and analytic domain leakage as a function
      of the iteration matrix spectral radius.} \textit{To test the
      behavior with respect to domain size, $n_i=50$, $n_i=100$,and
      $n_i=200$ were used. The red data was computed numerically by a
      domain-decomposed adjoint Neumann-Ulam implementation, the black
      fine-dashed data was generated by
      Eq~(\ref{eq:mean_chord_domain_leakage}), and the dashed-dotted
      black data was generated by
      Eq~(\ref{eq:wigner_domain_leakage}).}}
  \label{fig:measured_leakage}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{leakage_error.png}
  \end{center}
  \caption{\textbf{Measured and analytic domain leakage absolute error
      as a function of the iteration matrix spectral radius.}
    \textit{To test the behavior with respect to domain size, $n_i=5$
      (blue), $n_i=10$ (red), and $n_i=100$ (yellow) were used. The
      dashed lines represent the error using the Wigner approximation
      while the solid lines represent the error using the mean-chord
      approximation.}}
  \label{fig:measured_leakage}
\end{figure}
